\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-11-08 casestudies.tex
edits and tightening

Elimination of inductive types defined via \lsti{FInductive} is
mostly via the \lsti{FRecursion} and \lsti{FInduction} commands.
An exception is a handful of trivial ``inversion lemmas''.
For example, consider the lemma \lsti!\forall t, \neg step tm_true t!
stating that \lsti!tm_true! is irreducible.
If \lsti{step} were an ordinary inductive type, then it could be proved in Coq
simply by \lsti{intros t H; inversion H}.
But \lsti{step} is extensible. So one way to prove the lemma is
by \lsti{FInduction} on \lsti{step} and verifying that
a derived family does not accidentally make \lsti{tm_true} reducible.
%lemma still holds in derived families that add new constructors to \lsti{step}.
%
We observe that it is lighter-weight to use overriding (\cref{sec:override}) instead:
the programmer can specify that the proof of the lemma should be overridden
in any derived family that further binds \lsti{step},
and in return, they are permitted to treat \lsti{step} as an ordinary
inductive type in the proof and thus use \lsti{inversion} to prove it.
The plugin then automatically tries the same proof script in a derived
family to override the proof. Although proof scripts rather
than proof terms are reused, the practice seems justified
by the triviality of the lemmas and the terseness of the proof scripts.

>>>>

Elimination of inductive types defined via \lsti{FInductive} is
mostly via the \lsti{FRecursion} and \lsti{FInduction} commands.
An exception is a handful of trivial ``inversion lemmas''.
For example, consider the lemma \lsti!\forall t, \neg step tm_true t!
stating that \lsti!tm_true! is irreducible.
If \lsti{step} were an ordinary inductive type, then it could be proved in Coq
simply by \lsti{intros t H; inversion H}.
But \lsti{step} is extensible. So one way to prove the lemma is
by \lsti{FInduction} on \lsti{step} and verifying that
a derived family does not accidentally make \lsti{tm_true} reducible.
%lemma still holds in derived families that add new constructors to \lsti{step}.
%
We observe that it is lighter-weight to use overriding (\cref{sec:override}) instead:
the programmer can specify that the proof of the lemma should be overridden
in any derived family that further binds \lsti{step},
and in return, they are permitted to treat \lsti{step} as an ordinary
inductive type in the proof and thus use \lsti{inversion} to prove it.
The plugin then automatically tries the same proof script in a derived
family to override the proof. Although proof scripts rather
than proof terms are reused, this practice seems justified
by the triviality of the lemmas and the terseness of the proof scripts.

### 2022-11-08 coqimpl.tex
edits and tightening

For instance, consider field \lsti{env} in family \lsti{STLC}.
It is translated into a top-level module named \lsti{STLC\_env}.
This module has a parameter called \lsti{self} representing the enclosing
family: fields of the current family in the context of \lsti{env} can be
referenced through this \lsti{self}.
In particular, \lsti{env} is defined as \lsti{id -> option ty},
where \lsti{ty} is a late-bound reference to the \lsti{ty} field of the
enclosing family.
Hence, this reference to \lsti{ty} is translated in \lsti{STLC\_env} to \lsti{self.ty},
which is manifestly polymorphic to the enclosing family.
This translation of the \lsti{env} field can be shared
with a derived family even if it extends \lsti{ty} (e.g., \lsti{STLCProd})---%
no recompilation is needed because \lsti{self.ty} is not tied to a
particular definition of \lsti{ty}.

@@

Module \lsti{STLCFix} in \cref{fig:stlcfix-compiled} is generated upon
\lsti{End}~\lsti{STLCFix}, in the same fashion described above for \lsti{STLC}.
The translation makes code and proof sharing evident.
For example, the four case handlers (\lsti{subst\_tm_var} etc.) compiled for \lsti{STLC}
are reused to instantiate \lsti{subst}.
Translations of the definition of \lsti{env} and the proof of \lsti{typesafe},
are also reused to construct module \lsti{STLCFix}.

One may argue that because the first four constructors of \lsti{tm} are
repeated in \lsti{STLCFix}, the translation does not satisfy the
modular compilation requirement. Compared with alternatives that create
wrapper inductive types, we consider it a small price to pay in return
for the clarity of implementation especially given that compiled case handlers
are entirely reusable without rechecking.

>>>>

For instance,
consider field \lsti{env} in family \lsti{STLC}.
It is translated into a top-level module named \lsti{STLC\_env}.
This module has a parameter called \lsti{self} representing the enclosing
family: fields of the current family in the context of \lsti{env} can be
referenced through \lsti{self}.
In particular, \lsti{env} is defined as \lsti{id}\,->\,\lsti{option ty},
where \lsti{ty} is a late-bound reference to the \lsti{ty} field of the
enclosing family.
Hence, this reference to \lsti{ty} is translated in \lsti{STLC\_env} to \lsti{self.ty},
which is manifestly polymorphic to the enclosing family.
This translation of the \lsti{env} field can be shared
with a derived family even if it extends \lsti{ty} (e.g., \lsti{STLCProd})---%
no recompilation is needed because \lsti{self.ty} is not tied to a
particular definition of \lsti{ty}.

@@

Module \lsti{STLCFix} in \cref{fig:stlcfix-compiled} is generated upon
\lsti{End}~\lsti{STLCFix}, in the same fashion described above for \lsti{STLC}.
The translation makes code and proof sharing evident.
In particular, case handlers compiled for \lsti{STLC}
are reused to instantiate \lsti{subst}, \lsti{substlem}, and alike.
\lsti{STLC\_env} and \lsti{STLC\_typesafe}
are also reused to construct module \lsti{STLCFix}.
%
One may argue that because the first four constructors of \lsti{tm} are
repeated in \lsti{STLCFix}, the translation does not satisfy the
modular compilation requirement.
We could address this issue by using wrapper types,
but we consider restating constructors a small price to pay in return
for the clarity and concision of implementation.
We emphasize that compiled case handlers
are entirely reusable without rechecking, even with restated constructors.

### 2022-11-08 introduction.tex
edits and tightening

\item We contribute a prototypical implementation of our language
mechanism as a Coq plugin (\cref{sec:coqimpl}). The plugin works by
compiling surface-language terms into Gallina terms parameterized by
extensibility hooks.

\item We capture the essence of the new language mechanism formally by extending
Martin–Löf type theory with family polymorphism and extensible inductive
types (\cref{sec:metatheory2}).
We derive foundational meta\-theoretical results including consistency
and canonicity.
%\EDJ{I will suggest use ``foundational'' instead. Sterling at some places say canonicity is the most basic meta-theoretical property.}
We also formalize the translation implemented by the Coq plugin.

>>>>

\item We contribute a prototypical implementation of our language
mechanism as a Coq plugin (\cref{sec:coqimpl}). The plugin works by
compiling surface-language definitions into Coq modules parameterized by
explicit extensibility hooks.

\item We capture the essence of the new language mechanism formally by extending
Martin–Löf type theory with facilities to express family polymorphism (\cref{sec:metatheory2}).
We prove foundational meta\-theoretical results including consistency
and canonicity.
%\EDJ{I will suggest use ``foundational'' instead. Sterling at some places say canonicity is the most basic meta-theoretical property.}
%We also formalize the translation implemented by the Coq plugin.

### 2022-11-08 metatheory-condensed.tex
edits and tightening

\TT is intended as a foundational model. So unlike our programmer-facing plugin,
typing is structural rather than nominal, and
partial recursors and propositional equalities are not automated but have to
be explicitly defined.
\TT provides MLTT-style constructs that can be used to \emph{encode}
family polymorphism rather than building it in directly.
%
Most notably, \TT extends MLTT with what we call \emph{linkages}.\footnote{
Linkages are a namesake of the theoretical device through which prior work~\cite{zm2017} models
family polymorphism in an OO setting, but the technical details differ significantly.}
Linkages model families, so they are like tuples but with a twist: they
support late binding.
Unlike dependent tuples where a later component is \emph{existentially}
quantified over the earlier ones, a linkage component is \emph{universally}
quantified over---and thus polymorphic to---components preceding it.
%
\TT also features \emph{linkage transformers} that can be used to inductively
construct linkages from other linkages, by inheriting or overriding components of a linkage,
or by extending a linkage with new components.
%
Inductive types are modeled as W-types~\cite{martin1984intuitionistic} and their
extension as overriding.
%
Consistency and canonicity are proved by giving semantic models to the syntax.

@@

The third premise
$\goodTerm{\Gm,\TyTkg{\lsig}}{\seal}{\sub{A}{\SubstWeak{1}}}$
in rules \ruleref{l/add} and \ruleref{lsig/add} is
responsible for hiding W-type signatures.
Here, $\TyTkg{\lsig}$ is a dependent tuple type (rule \ruleref{tyeq/pk/add})
that packages the types of all fields preceding the current field~$t$,
and $\seal$ is the term turning the tuple $\Tkg{\lkg}$ of type
$\TyTkg{\lsig}$ (rule \ruleref{tmeq/pk/add}) into a new tuple of type~$A$
that hides W-type signatures behind $\cU$, if necessary.
%
It is straightforward to find the~$\seal$ that fits the bill, though this process
is not automated in \TT.
In particular, when no hiding is needed, $\seal$ is $\var{0}$ (the identity) and $A$ is
$\TyTkg{\lsig}$.
In \cref{fig:stlc-linkage-typing}, $\seal_1$, $\seal_2$, and $\seal_{10}$ are identities,
and $\seal_6$ turns
\begingroup
\fontsize{9}{9}\selectfont%
$\big[\texttt{tm} : \TyS{\wcode{\wsig_{\texttt{tm}}}},\ \texttt{tm\_unit} : \El{\texttt{tm}},\ \dadada\big]$
\endgroup
into
\begingroup
\fontsize{9}{9}\selectfont%
$\big[\texttt{tm} : \cU,\ \texttt{tm\_unit} : \El{\texttt{tm}},\ \dadada\big]$%
\endgroup
.

When a family is completed (e.g., \lsti!End!~\lsti!STLC!),
a linkage~$\lkg$ containing all the fields is available
(e.g., $\lkg_{53}$ in \cref{fig:stlc-linkage-typing}).
Fields of the family can then be accessed by projecting them out of the tuple
$\Tkg{\lkg}$.

>>>>

\TT is intended as a foundational model. So unlike our programmer-facing plugin,
typing is structural rather than nominal, and
partial recursors and propositional equalities are not automated but have to
be explicitly defined.
\TT provides MLTT-style constructs that can be used to \emph{encode}
family polymorphism rather than building it in directly.
%
Most notably, \TT extends MLTT with what we call \emph{linkages}.\footnote{
Linkages are a namesake of the theoretical device through which prior work~\cite{zm2017} models
family polymorphism in an OO setting, but the technical details differ significantly.}
Linkages model families, so they are tuples composed of fields but with a twist: they
support late binding.
Unlike dependent tuples where a later component is \emph{existentially}
quantified over the earlier ones, a linkage component is \emph{universally}
quantified over---and thus polymorphic to---components preceding it.
%
\TT also features \emph{linkage transformers} that can be used to inductively
construct linkages from other linkages, by inheriting or overriding components of a linkage,
or by extending a linkage with new components.
%
Inductive types are modeled as W-types~\cite{martin1984intuitionistic} and their
extension as overriding.
%
Consistency and canonicity are proved by giving semantic models to the syntax.

@@

The third premise
$\goodTerm{\Gm,\TyTkg{\lsig}}{\seal}{\sub{A}{\SubstWeak{1}}}$
in rules \ruleref{l/add} and \ruleref{lsig/add} is
responsible for hiding W-type signatures.
Here, $\TyTkg{\lsig}$ is a dependent tuple type (rule \ruleref{tyeq/pk/add})
that packages the types of all fields preceding the current field~$t$,
and $\seal$ is the term turning the tuple $\Tkg{\lkg}$ of type
$\TyTkg{\lsig}$ (rule \ruleref{tmeq/pk/add}) into a new tuple of type~$A$
that hides W-type signatures behind $\cU$, if necessary.
%
It is straightforward to find the~$\seal$ that fits the bill, though this process
is not automated in \TT.
In particular, when no hiding is needed, $\seal$ is $\var{0}$ (the identity) and $A$ is
$\TyTkg{\lsig}$.
In \cref{fig:stlc-linkage-typing}, $\seal_1$, $\seal_2$, and $\seal_{10}$ are identities,
and $\seal_6$ is from
\begingroup
\fontsize{9}{9}\selectfont%
$\big[\texttt{tm} : \TyS{\wcode{\wsig_{\texttt{tm}}}},\ \texttt{tm\_unit} : \El{\texttt{tm}},\ \dadada\big]$
\endgroup
to
\begingroup
\fontsize{9}{9}\selectfont%
$\big[\texttt{tm} : \cU,\ \texttt{tm\_unit} : \El{\texttt{tm}},\ \dadada\big]$%
\endgroup
.

When a family is concluded (e.g., \lsti!End!~\lsti!STLC!),
a linkage~$\lkg$ containing all the fields is available
(e.g., $\lkg_{53}$ in \cref{fig:stlc-linkage-typing}).
Fields of the family can then be accessed by projecting them out of the tuple
$\Tkg{\lkg}$.
